---
title: "MS-based proteomics for *R. eutropha*"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook: 
    theme: spacelab
    toc: yes
---


## Description

This R notebook is a bioinformatics pipeline to **process and analyze MS based peptide/protein abundance data** for the chemolithoautotroph *Ralstonia eutropha* (a.k.a. *Cupriavidus necator*).

Proteomics data was obtained using the following work flow (to be added...).


## Libraries

```{r, message = FALSE}
# loading libraries
library(lattice)
library(latticeExtra)
library(latticetools)
library(tidyverse)
library(stringi)
```

## Data import

Define the data source directories. Some of them are external in the sense of not included in the accompanying data folder of this R notebook. The main proteomics data is loaded from the R ShinyProt directory that can also be found on [github](https://github.com/m-jahn/ShinyProt) and interactively browsed and searched in the [ShinyProt web app](https://m-jahn.shinyapps.io/ShinyProt/).

```{r, message = FALSE}
Reutropha_proteomics <- "~/Documents/SciLifeLab/Resources/R_projects/ShinyProt/data/Ralstonia_eutropha.Rdata"
load(Reutropha_proteomics)
```


## Overview on detected peptides and proteins

### Total number of quantified proteins and missing proteome

In total, the following number of proteins was quantified, out of theoretical total of 6,614 proteins ([Uniprot reference genome](https://www.uniprot.org/uniprot/?query=organism%3A%22Cupriavidus+necator+%28strain+ATCC+17699+%2F+H16+%2F+DSM+428+%2F+Stanier+337%29+%28Ralstonia+eutropha%29+%5B381666%5D%22+AND+proteome%3Aup000008210&sort=score), July 31 2020). That represents roughly 81 % by number and much more by mass (see below for estimation)

```{r}
n_quantified_prot <- Ralstonia_eutropha %>% pull(uniprot) %>% unique %>% length
print(n_quantified_prot)
print(n_quantified_prot/6614*100)
```

We can also estimate the coverage in terms of protein mass, by assuming that the 1257 missing proteins are of average mass or lower than average mass than the detected proteins. The following simple calculation simulates missing protein mass with an average abundance of the lower quantile of detected proteins. For this purpose we simply pick a standard condition, such as growth on fructose.

```{r}
quantified_protein <- Ralstonia_eutropha %>%
  
  # pick a certain condition
  filter(substrate == "fructose", growthrate == 0.25) %>%
  pull(mean_intensity)

# determine quantiles of raw quantification intensity
quantified_protein %>% quantile(na.rm = TRUE)
```

Now we just simulate that the 1257 non-detected proteins have an average mass similar to that of the protein with 25% lowest abundance. The missing protein abundance will then sum up to less 1% of the total estimated proteome, meaning we can detect more than 99% of the proteome by mass.

```{r}
missing_protein <- 1257 * quantile(na.rm = TRUE, quantified_protein)[2] %>% as.numeric()
missing_protein_percent <- missing_protein/(missing_protein + sum(quantified_protein, na.rm = TRUE))
paste("missing protein in % total mass:", round(missing_protein_percent*100, 3))
```

### Number of quantified peptides per protein

```{r, fig.width = 5, fig.height = 2.5}
plot_quant_pep <- xyplot(sort(n_peptides, decreasing = TRUE) ~ 
      1:length(protein),
    filter(Ralstonia_eutropha, substrate == "fructose", growthrate == 0.25),
    xlab = "protein", ylab = "n peptides",
    par.settings = custom.lattice, 
    ylim = c(0, 80), xlim = c(0, 5500),
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.barplot(x, y, col = NA, fill = grey(0.8), ewidth = 0.6)
      xhalf = length(unique(x))/2
      panel.lines(x = c(0, xhalf, xhalf), y = c(y[xhalf], y[xhalf], 0), col = 1)
      panel.text(x = 10, y = y[xhalf]*3, pos= 4, cex = 0.8, 
        labels = paste0(round(xhalf), " proteins with >= ", y[xhalf], " peptides"))
    }
  )

print(plot_quant_pep)
```

### Number of quantified peptides per protein, inverted

```{r, fig.width = 5, fig.height = 2.5}
plot_quant_pep_2 <-  Ralstonia_eutropha %>%
  
  # rearrange n_peptides to have another type of overview
  filter(substrate == "fructose", growthrate == 0.25) %>%
  pull(n_peptides) %>% table %>% as_tibble %>%
      rename(., pep = `.`, prot = n) %>% mutate(pep = as.numeric(pep)) %>%
  
  # plot
  xyplot(prot ~ pep, .,
    xlab = "n peptides", ylab = "n proteins", ylim = c(-5, 1155),
    par.settings = custom.lattice, xlim = c(0, 80),
    panel = function(x, y,...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.barchart(x, y, horizontal = FALSE, box.width = 1,
        border = NA, col = grey(0.8), ...)
    }
  )

print(plot_quant_pep_2)
```


### Number of protein quantifications per replicate

```{r, fig.width = 5, fig.height = 2.5, message = FALSE}
plot_quant_prot1 <- Ralstonia_eutropha %>%
  
  # protein quantifications per replicate
  gather(replicate, raw_intensity, R1:R4) %>%
  group_by(substrate, growthrate, replicate) %>%
  summarize(quant_proteins = sum(!is.na(raw_intensity))) %>%
  
  # and plot
  xyplot(quant_proteins ~ 1:length(quant_proteins), .,
    xlab = "sample", ylab = "quantified proteins",
    par.settings = custom.lattice, 
    ylim = c(0, 5500), xlim = c(0, 81),
      panel = function(x, y,...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.barplot(x, y, col = NA, fill = grey(0.8), ewidth = 0.5)
    }
  )

print(plot_quant_prot1)
```

### Number of proteins quantified in every run


```{r, fig.width = 5, fig.height = 2.5, message = FALSE}
plot_quant_prot2 <- Ralstonia_eutropha %>%
  
  # protein quantifications per replicate
  gather(replicate, raw_intensity, R1:R4) %>%
  group_by(protein) %>%
  summarize(quant_in_runs = sum(!is.na(raw_intensity))) %>%
  pull(quant_in_runs) %>% table %>% enframe %>%
  
  # and plot
  xyplot(value ~ factor(name), .,
    xlab = "in number of runs", 
    ylab = "quantified proteins",
    par.settings = custom.lattice, 
    ylim = c(0, 3000), xlim = c(0, 79),
      panel = function(x, y,...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.barplot(x, y, col = NA, fill = grey(0.8), ewidth = 0.5)
      panel.abline(v = 70, col = 1)
      panel.text(x = 25, y = 1500, pos= 4, cex = 0.8, 
        labels = paste0(sum(y[70:79]), " proteins quantified\nin > 70 out of 80 runs"))
    }
  )

print(plot_quant_prot2)
```


## Sample overview and quality control


### Raw intensity per sample and replicate

Raw intensity here is the dimensionless MS 'intensity', that means the quantified area under the curve of MS1 spectra for peptides, summed up per protein. One replicate that was missing for condition Fructose, growth rate 0.1, R2, was temporarily replaced by R1 for this plot, because densityplot was otherwise giving an error message (because of missing values).

```{r, message = FALSE}
densityplot(~ log10(R1) + log10(R2) + log10(R3) + log10(R4) | condition, 
  Ralstonia_eutropha %>% mutate(R2 = case_when(
    condition == "FRC 0.1" ~ R1, TRUE ~ R2)), 
  auto.key = list(columns = 4), layout = c(5, 4), 
  par.settings = custom.colorblind(), xlab = "log10 intensity",
  scales = list(alternating = FALSE), as.table = TRUE,
  panel = function(x, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.superpose(x, ...)
  },
  panel.groups = function(x, ...) {
    panel.densityplot(x, plot.points = FALSE, ...)
    panel.abline(v = median(x, na.rm = TRUE), lty = 2, col = grey(0.5))
  }
)
```

### Variation per sample and replicate

Log 10 median intensity versus log 10 CV.

```{r, fig.width = 8, fig.height = 6}
library(hexbin)
hexbinplot(log10(CV) ~ log10(median_intensity) | condition, 
  Ralstonia_eutropha,
  layout = c(5, 4), 
  par.settings = custom.colorblind(),
  scales = list(alternating = FALSE), aspect = 0.9,
  ylim = c(-3, 0.5),
  colramp = colorRampPalette(custom.lattice()$superpose.polygon$col[3:1])
)
```

A densityplot of the coeffcient of variation (CV) for the four replicates per protein, broken down by sample.
This result shows that the variation is considerably higher than an 'ideal' MS-based proteomics experiment, where average CV can be as low as 10%. Here, variation is high as 50%.

```{r, fig.width = 8, fig.height = 6}
densityplot(~ CV | condition,
  Ralstonia_eutropha,
  as.table = TRUE, lwd = 2, 
  par.settings = custom.colorblind(), pch = ".",
  panel = function(x, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.densityplot(x, ...)
    panel.ablineq(v = median(x, na.rm = TRUE), adj = -0.1,
      lty = 2, col = grey(0.5), fontfamily = "FreeSans")
  }
)
```
### Similarity between replicates from identifications

We can determine the overall similarity of samples (replicates and conditions) towards each other. A simple strategy for this is to use **PCA** or **nMDS**, the latter is an iterative and therefore not fully deterministic approach as it depends on the starting coordinates. However it is useful to compare how different samples or replicates 'cluster' together. Two replicates need to be removed from the data before, one corrupt sample that is missing (`FRC_0.1_R2`), and one that is an outlier (`NLIM_0.05_R1`).

The strategy for nMDS is to reshape all measurements per sample into a matrix and compute the 'distance' by a default measure. nMDS then tries to arrange each sample as a dot on a plane, taking optimal the distance to its neighbors into account. This approach might not give a perfect result and may contain contradictions indicated by the `stress` level.

```{r, fig.width = 10, fig.height = 4, message = FALSE}
# load required libraries
library(dendextend)
library(vegan)

# first need to rearrange raw data so that we obtain a 'wide' table/matrix
dist_mat <- Ralstonia_eutropha %>% select(uniprot, condition, R1:R4) %>%
  gather(replicate, intensity, R1:R4) %>%
  unite(condition, condition, replicate) %>%
  spread(condition, intensity) %>% 
  
  # remove missing/outlier samples and coerce to matrix
  select(-uniprot, -all_of(c("FRC 0.1_R2", "NLIM 0.05_R1"))) %>%
  as.matrix %>% t

# plot sample similarity as dendrogram
plot_cols <- custom.colorblind()$superpose.polygon$col[1:4]
cluster <- hclust(dist(dist_mat), method = "ward.D2")
plot(color_branches(cluster, col = rep(plot_cols, each = 20)[-c(26, 41)][cluster$order]))
```

We can see that many replicates cluster nicely together, samples also cluster predominantly by carbon/nitrogen limitation. For example, many samples from formic acid and nitrogen limitation cluster on the left side, and many samples from fructose cluster on the right side.

```{r, fig.height = 4, fig.width = 4, message = FALSE}
library(tactile)

# run nMDS analysis
NMDS <- dist_mat %>% dist %>% metaMDS

# and plot result
df_nmds <- NMDS$points %>% as_tibble(rownames = "condition") %>%
  separate(condition, into = c("condition", "growth_rate", "replicate"), sep = "[ _]") %>%
  mutate(across(matches("MDS[12]"), function(x) x/10^11)) %>%
  mutate(growth_rate = as.numeric(growth_rate)*10)

xyplot(MDS2 ~ MDS1, df_nmds,
  groups = condition,
  pch = 19, size = df_nmds$growth_rate,
  par.settings = custom.colorblind(),
  panel = function(x, y, size, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.bubbleplot(x, y, z = size+0.5, col = grey(0.9), ...)
    panel.bubbleplot(x, y, z = size, ...)
    panel.key(..., cex = 0.7, corner = c(0.95, 0.05))
    for (p in 1:5) {panel.key(labels = as.character(p/20), col = grey(0.7),
      pch = 19, cex = 0.7, point.cex = p/2, corner = c(0.95, 0.95-(0.09*p)^1.5))
    }
  }
)
```

As an alternative, we can also use principal component analysis, PCA, and see if this is similarly conclusive than NMDS. PCA reduces the dimensionaity of data by using a sort of projection on a 2D plane. In contrast to NMDS it is a determinstic and non-iterative procedure. It computes principal components (PC) that have a certain proportion of explanatory power of the variation between data points.

The result shows a clustering that is less compelling as the NMDS but with similar message: Fructose and succinate are somewhat closer related, for the low growth rates also N-limitation. In the NMDS, formic acid is very far apart from the other samples while in the PCA this trend is not that clear. There seems to be trend of low and higher growth rates clustering apart from each other, which is more clear in NMDS.

```{r, fig.height = 4, fig.width = 4, message = FALSE}
PCA_result <- dist_mat %>% t %>% na.omit %>%
  
  # scale and perform PCA analysis. One can also log10-transform expression data to get a 
  # comparison that is stronger between relative instead of absolute differences
  # in protein abundance
  scale %>% prcomp(center = TRUE) 
  
# summary of variation
summary(PCA_result)$importance[, 1:6]
  
# plot PC1, PC2 and PC3
PCA_result$rotation %>% as_tibble %>%
  mutate(
    condition = row.names(dist_mat) %>% stri_extract_all_regex("^[A-Z]*") %>% unlist,
    growth_rate = row.names(dist_mat) %>% stri_extract_all_regex("0.[0-9]*") %>% unlist
  ) %>%
  
  xyplot(PC2 ~ PC1, .,
    par.settings = custom.colorblind(), groups = condition,
    pch = 19, size = 10*as.numeric(.[["growth_rate"]]),
    panel = function(x, y, size, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.bubbleplot(x, y, z = size, ...)
      panel.key(..., cex = 0.7, corner = c(0.05, 0.05))
      for (p in 1:5) {panel.key(labels = as.character(p/20), col = grey(0.7),
        pch = 19, cex = 0.7, point.cex = p/2, corner = c(0.05, 0.95-(0.09*p)^1.5))
      }
    }
  )
```

