---
title: "Physiological measurements from *R. eutropha* chemostats"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook: 
    theme: spacelab
    toc: yes
---

## Description

This R notebook contains data and analysis regarding physiological measurements obtained from chemostat cultivations of *Ralstonia eutropha* (a.k.a. *Cupriavidus necator*).

## Libraries

```{r, messages = FALSE}
library(lattice)
library(latticeExtra)
library(latticetools)
library(tidyverse)
```

## Chemostat cultivation, biomass and growth rate

### Import and reshape cultivation data

First we load the three cultivation data tables. These contain all four limiting conditions. We are thinning out the OD measurement data by 75% so that file sizes become smaller and more plottable.

```{r}
# load chemostat measurements
df_chem <- read.csv("../data/input/20180921_chemostat_OD.csv") %>%
  
  # remove outlier data points originating from sampling
  filter(!od_value > 0.6) %>%
  
  # thinning data to reduce file size
  slice_sample(n = 4000)
```

### OD and growth rate

The next step is to plot the OD<sub>720</sub> and growth rate for all cultivations.

```{r, fig.width = 5, fig.height = 5.5}
plot_OD <- xyplot(od_value ~ batchtime_h | condition, 
  arrange(df_chem, batchtime_h),
  groups = replicate, par.settings = custom.colorblind(),
  as.table = TRUE,, between = list(x = 0.5, y = 0.5), type = "l", lwd = 2,
  ylim = c(-0.06, 0.86), ylab = expression("OD"[720*nm])
)

plot_dil <- xyplot(dilution_rate ~ batchtime_h | condition, 
  filter(df_chem, replicate == 1), 
  between = list(x = 0.5, y = 0.5),
  par.settings = custom.colorblind(), 
  layout = c(1,4), as.table = TRUE,
  type = "l", xlim = c(0, 335), ylim = c(-0.03, 0.43),
  xlab = "time [h]", ylab = expression("µ [h"^-1*"]"),
  panel = function(x, y, ...){
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.horizonplot(x, y, col.regions = grey(0.7, alpha = 0.5))
  }
)

print(doubleYScale(plot_dil, plot_OD, use.style = FALSE, add.ylab2 = TRUE))
```

### Calculate biomass yield and maintenance


First we summarize average OD and growth rate for all cultivations leaving out the first 25 h batch growth phase.

```{r}
df_summary <- filter(df_chem, batchtime_h > 25) %>%
  
  # summarize average OD
  group_by(condition, dilution_rate, replicate) %>%
  summarize(OD720 = mean(od_value, na.rm = TRUE)) %>%
  
  # then we need medium substrate concentration for FA, FRC, NH4CL, SUC
  mutate(c_substrate = case_when(
    condition == "ammonium" ~ 0.025,
    condition == "formate" ~ 1.0,
    TRUE ~ 0.5
  ))
```

For yield calculations, the residual substrate (mg) needs to be subtracted from feed substrate concentration (g) to determine portion of substrate that is taken up.

```{r, message = FALSE}
df_residual <- read_csv("../data/input/20181116_chemostat_residual_substrate.csv") %>%
  rename(condition = limitation) %>%
  mutate(condition = recode(condition, `formic acid` = "formate", `Nlim - fructose` = "ammonium")) %>%
  mutate(c_residual_g_L = mean_concentration_mg.L/1000 %>% replace(., . < 0, 0))

# join the summary df with residual substrate data
df_summary <- left_join(df_summary, df_residual %>%
    select(condition, dilution_rate, c_residual_g_L) %>%
    filter(!duplicated(c_residual_g_L), condition != "ammonium")) %>%
  mutate(c_residual_g_L = c_residual_g_L %>% replace_na(0))

print(head(df_summary))
```

Nest step is to add the DCW in g/L depending on OD and condition. The DCW-OD relationship for each limiting condition was determined in a different experiment (calibration performed 2020-07-30). 50 mL shake flasks were used to determine OD at 720 nm, final gDCW per OD<sub>720</sub> and L, and maximum growth rate (in duplicates). The gDCW per OD was similar for all C-limiting condition and more different for the N-limitation due to accumulation of PHB granules. They absorb/scatter stronger than cellular biomass and therefore lead to over-estimation of cell density/biomass.


We calculate substrate uptake rate q_S with the corrected substrate concentration (S_uptake = S_feed - S_residual). The rate q_S is then dilution rate times feed concentration divided by biomass in the culture: qS [g/OD720*L] = D [h-1] * c [g/L] / OD720. All other following calculations are just variations of that converting OD to gDCW per L, or converting g to C/N-mole.
  
```{r}
df_summary <- df_summary %>% mutate(gDCW_L = case_when(
    condition == "fructose" ~ OD720*0.3980,
    condition == "succinate" ~ OD720*0.3392,
    condition == "formate" ~ OD720*0.3600,
    condition == "ammonium" ~ OD720*0.2766
  )) %>%
  
  # substrate uptake rate per OD
  # qS [g/OD720*L] = D [h-1] * c [g/L] / OD720
  mutate(qS_g_OD720 = dilution_rate*(c_substrate-c_residual_g_L)/OD720) %>%
  
  # substrate uptake rate per gDCW
  # qS [g/gDCW] = D [h-1] * c [g/L] / gDCW * L
  mutate(qS_g_gDCW = dilution_rate*(c_substrate-c_residual_g_L)/gDCW_L) %>%
  
  # substrate uptake rate in C-mol per C-mol biomass (25.35 gDCW/Cmol = 0.03944 Cmol/gDCW)
  # (source: Grunwald et al., Mirco Biotech, 2015)
  mutate(Cmol_g = case_when(
    condition == "formate" ~ 1/46.03 * 1,
    condition == "fructose" ~ 1/180.16 * 6,
    condition == "succinate" ~ 1/118.09 * 4,
    condition == "ammonium" ~ 1/53.49 * 1 # here it is N-mol! NH4Cl has 1 N
  )) %>%
  
  # for NH4Cl, we actually use N-mol per N-mol biomass per h. 
  # N-mol in biomass is around 4x lower than C mol (bionumbers)
  mutate(qS_Cmol_Cmol_DCW = case_when(
    condition == "ammonium" ~ qS_g_gDCW * Cmol_g / (0.03944/4),
    TRUE ~ qS_g_gDCW * Cmol_g / 0.03944
  ))

head(df_summary)
```

Finally we can create a so-called Herbert-Pirt plot, that is growth rate versus substrate uptake rate. This plot will usually reveal a linear relationship, meaning that substrate uptake and growth are proportional to each other. If there is a kink (change in slope) of the data and one linear model would not be sufficient to describe the data, then the yield has changed, i.e. cells start using a different pathway for energy generation (e.g. respiration versus fermentation). In the default scenario (linear relationship), the yield is the slope of the linear model and the maintenance coeffcient is the intersection with the Y axis.

```{r, fig.width = 6.7, fig.height = 6.7}
HP_plot <- function(data, xlimits = c(0, 0.3), ylimits = c(0, 1)) {
  xyplot(qS_g_gDCW ~ dilution_rate | condition, data,
    par.settings = custom.lattice, lwd = 1.5,
    xlim = xlimits, ylim = ylimits,
    ylab = expression("q"[S]*" [g h"^-1*" gDCW"^-1*"]"),
    xlab = expression('µ [h'^'-1'*']'),
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.xyplot(x, y, cex = 0.9, ...)
      # regression line through linear part of the data (omit µ = 0.25/h)
      sel <- x != 0.25; x = x[sel]; y = y[sel]
      panel.lmlineq(x, y, fontfamily = "FreeSans", 
        pos = 3, offset = 7, r.squared = TRUE, cex = 0.7, col.text = grey(0.3), ...)
      coef <- lm(y ~ x, data.frame(x, y))$coeff
      # displaying maintenance and yield coefficients
      panel.abline(h = coef[[1]], lty = 2, ...)
      panel.text(0.15, coef[[1]],
        paste("ms =", round(coef[[1]], 3), "g h-1 gDCW-1"),
        col = grey(0.3), pos = 3, cex = 0.7)
      panel.text(0.15, coef[[1]], paste(expression("Yx/S ="),
        round(1/coef[[2]], 3), "gDCW g-1"),
        col = grey(0.3), pos = 1, cex = 0.7)
    }
  )
}

print(HP_plot(data = filter(df_summary, condition == "formate"),
  ylimits = c(-7/4, 7)), split = c(1,1,2,2), more = TRUE)
print(HP_plot(data = filter(df_summary, condition == "fructose"),
  ylimits = c(-1.5/4, 1.5)), split = c(2,1,2,2), more = TRUE)
print(HP_plot(data = filter(df_summary, condition == "ammonium"),
  ylimits = c(-0.25/4, 0.25)), split = c(1,2,2,2), more = TRUE)
print(HP_plot(data = filter(df_summary, condition == "succinate"),
  ylimits = c(-1.5/4, 1.5)), split = c(2,2,2,2))
```


```{r, include = FALSE}
# silently export figure
svg("../figures/figure_growth_vs_uptake.svg", width = 6.7, height = 6.7)
print(HP_plot(data = filter(df_summary, condition == "formate"),
  ylimits = c(-7/4, 7)), split = c(1,1,2,2), more = TRUE)
print(HP_plot(data = filter(df_summary, condition=="fructose"),
  ylimits = c(-1.5/4, 1.5)), split = c(2,1,2,2), more = TRUE)
print(HP_plot(data = filter(df_summary, condition=="ammonium"),
  ylimits = c(-0.25/4, 0.25)), split = c(1,2,2,2), more = TRUE)
print(HP_plot(data = filter(df_summary, condition=="succinate"),
  ylimits = c(-1.5/4, 1.5)), split = c(2,2,2,2))
dev.off()
```


A similar plot could be made for C-mole and N-mole per C-mole biomass instead of [gram] units. Since the data is similar and only the units change, we just extract the slope and intercept of the four linear models for each substrate.

```{r, message = FALSE}
get_lm_coeff <- function(x, y) {
  linmod <- lm(y ~ x, data.frame(x = x, y = y))
  c(round(linmod$coeff[[1]], 3), round(1/linmod$coeff[[2]], 3))
}
df_summary %>% group_by(condition) %>%
  summarize(value = get_lm_coeff(dilution_rate, qS_Cmol_Cmol_DCW)) %>% 
  ungroup %>%
  mutate(variable = rep(c("mS [Cmole Cmole_DCW-1 h-1]", "Yx/S [Cmole DCW Cmole-1]"), 4))
```



### Biomass, residual substrate

This is just an auxiliary plot for the cultivation overview of OD and growth rate from above. In this plot, the average biomass concentration is shown per condition.

```{r, fig.width = 6, fig.height =2.5}
xyplot(gDCW_L ~ factor(dilution_rate) | condition, df_summary,
  as.table = FALSE, xlab = "µ [h-1]", ylab = "biomass [gDCW/L]",
  between = list(x = 0.7, y = 0.7),
  scales = list(alternating = FALSE), lwd = 3,
  par.settings = custom.colorblind(), layout = c(4, 1),
  pch = 19, ylim = c(-0.02, 0.27),
  panel = function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.errbars(x, y, ewidth = 0, ...)
  }
)
```
Finally for this section we can plot raw measurements using double Y scale (% and ppm)

```{r, fig.width = 6, fig.height = 2.5}
plot_residual <- xyplot(concentration_mg.L ~ factor(dilution_rate) | condition,
  df_residual,
  par.settings = custom.colorblind(), as.table=TRUE,
  pch = 19, ylim = c(-50, 1450), layout = c(4, 1),
  between = list(x = 0.5, y = 0.5),
  scales = list(alternating = FALSE, x = list(at = c(1,3,5))),
  xlab = expression("µ [h"^-1*"]"), ylab = "concentration [mg/L]",
  panel = function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.barplot(x, y, ewidth = 0.3, lwd = 2, ...)
    panel.xyplot(x, y, alpha = 0.3, ...)
  }
)

print(plot_residual)
```


```{r, include = FALSE}
# silently export figure
svg("../figures/figure_chemostat_cult.svg", width = 5, height = 7.5)
print(doubleYScale(plot_dil, plot_OD, use.style = FALSE, add.ylab2 = TRUE), 
  position = c(0,0.27,1,1), more = TRUE)
print(plot_residual, position = c(-0.01,0,1,0.32))
dev.off()
```


## PHB content of cells

The following section analyzes data obtained from chemostat cultivations of the *R. eutropha* wild type (WT, PHB+) and a PHB knockout mutant (PHB-) obtained by random mutagenesis. Both strains were cultivated in chemostats with stepwise increase of dilutiuon rate (equals growth rate). Different parameters of these cultivations were determined among, those poly-hydroxybutyrate (PHB) content. PHB content was determined using Nile red staining (relative, not absolute quantification).

First we import PHB quantification data per growth rate.

```{r, message = FALSE}
df <- read_csv("../data/input/20190716_chemostat_PHB.csv")
```

Next we reshape and filter the data a bit.

```{r}
df <- filter(df, mu <= 0.3) %>%
  mutate(condition = recode(
    condition, "C-limitation" = "fructose", "N-limitation" = "ammonium")) %>%
  
  mutate(replicate = if_else(channel > 4, channel-4, channel))

head(df)
```

Then we plot relative PHB content over growth rate. In a second step we *estimate* absolute PHB content by using data from the literature as help. We can set the maximum observed Nile Red fluorescence to the average relative PHB content obtained under N limitation as reported in the literature. For example Steinbuechel & Schlegel, AMB, 1989 reported a PHB content of 0.4 g/g protein under N-limitation. With 0.65 g protein/gDCW, that equals 0.4 x 0.65 = 0.26 g/gDCW PHB. A second reference, Wu *et al*., Kor J Chem Eng, 2009, determined that up to 41.8% biomass is PHB (= 0.41 g/gDCW).

```{r, fig.width = 5, fig.height = 3}
plot_NR_fluorescence <- xyplot(I_nr/OD720_nr/10^6 ~ factor(mu) | condition,
  df, groups = strain,
  par.settings = custom.colorblind(), 
  pch = 19, lwd = 2,
  as.table = TRUE, between = list(x = 0.5, y = 0.5),
  xlab = expression("µ [h"^-1*"]"), 
  ylab = expression("FI"["nile red"]*" OD"[720]*""^-1),
  scales = list(alternating = FALSE),
  panel=function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.errbars(x, y, ewidth = 0, ...)
    panel.key(..., which.panel = 2, corner = c(0.9, 0.9))
  }
)


df <- df %>% mutate(
  rel_PHB_g_gDCW = I_nr/OD720_nr/max(I_nr/OD720_nr)*0.41
)

plot_rel_PH_content <- xyplot(rel_PHB_g_gDCW ~ factor(mu) | condition,
  df, groups = strain,
  ylab = expression("PHB [g gDCW"^-1*"]"),
  panel=function(x, y, ...) {
    panel.errbars(x, y, ewidth = 0, ...)
  }
)

doubleYScale(plot_NR_fluorescence,
  plot_rel_PH_content, use.style = FALSE, add.ylab2 = TRUE) %>% print
```

```{r, include = FALSE}
# silently export svg figures
svg("../figures/figure_PHB_content.svg", width = 5, height = 3)
doubleYScale(plot_NR_fluorescence,
  plot_rel_PH_content, use.style = FALSE, add.ylab2 = TRUE) %>% print
dev.off()
```

## O<sub>2</sub> consumption and CO<sub>2</sub> emission

Outlet gas composition was analyzed with specific optical sensors for O<sub>2</sub> and CO<sub>2</sub>.
The gas uptake rate can then be determined knowing the input concentration, output concentration, flow rate per reactor, and biomass concentration per reactor. First load raw measurement data.

```{r, message = FALSE}
df_gas <- read_csv("../data/input/20190728_chemostat_gas.csv")
```

Next we summarize all measurements to one measurement per condition and replicate. Then we use measurements for the replicates to calculate the mean gas uptake and emission rate normalized to biomass concentration.

```{r}
df_gas <- df_gas %>%
  
  # remove mu higher than 0.3
  filter(mu <= 0.3) %>%
  
  # determine average delta air flow per cond and replicate
  group_by(condition, strain, mu, type, channel, `flow_rate [ml/min]`) %>%
  summarize(flow_mL_min = mean(`flow [ml/min]`)) %>%
  
  # renumber replicates from 1 to n
  group_by(condition, strain, mu, type) %>%
  mutate(replicate = seq_along(channel)) %>%
  arrange(condition, strain, mu, type, replicate) %>%
  
  # obtain biomass concentration from first table
  ungroup %>% left_join(
    group_by(df, strain, condition, mu) %>%
    summarize(gDCW_mL = mean(`DCW [g/ml]`, na.rm = TRUE)) %>% 
    ungroup
    )
```

Now that we have the biomass concentration in gDCW/mL, we can determine the CO<sub>2</sub> emission and O<sub>2</sub>. The culture volume is 65 mL and the mass per volume CO<sub>2</sub> is 1.842 mg/mL (air at 1 atm and 20 °C). The mass per volume O<sub>2</sub> is 1.314 mg/mL (air at 1 atm and 20 °C)
The CO<sub>2</sub> emission rate can then be calculated as:

qCO2 [mL min^-1 gDCW^-1] = flow_CO2 [mL min^-1] / (biomass gDCW mL^-1 * V_culture mL)
qCO2 [g h^-1 gDCW^-1] = 


```{r}
df_gas <- df_gas %>% 
  
  # add volume to mass conversion
  mutate(m_gas_mg_mL = if_else(type == "CO2", 1.842, 1.314)) %>%
  
  mutate(
  q_mL_min_gDCW = flow_mL_min / (gDCW_mL * 65),
  q_g_h_gDCW = q_mL_min_gDCW * m_gas_mg_mL * 60 / 1000
  )

head(df_gas)
```



```{r, fig.width = 5, fig.height = 5.5}
plot_CO2_em <- xyplot(q_g_h_gDCW ~ factor(mu) | condition,
  filter(df_gas, type == "CO2"), groups = strain,
  par.settings = custom.colorblind(), 
  pch = 19, lwd = 2, ylim = c(0, 1.5),
  as.table = TRUE, between = list(x = 0.5, y = 0.5),
  xlab = expression("µ [h"^-1*"]"), 
  ylab = expression("q"[CO2]*" g gDCW"^-1*" h"^-1),
  scales = list(alternating = FALSE),
  panel=function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.errbars(x, y, ewidth = 0, ...)
    panel.key(..., which.panel = 2, corner = c(0.9, 0.9))
  }
)

plot_O2_up <- xyplot(q_g_h_gDCW*-1 ~ factor(mu) | condition,
  filter(df_gas, type == "O2"), groups = strain,
  par.settings = custom.colorblind(), 
  pch = 19, lwd = 2, ylim = c(0, 1.5),
  as.table = TRUE, between = list(x = 0.5, y = 0.5),
  xlab = expression("µ [h"^-1*"]"), 
  ylab = expression("q"[O2]*" g gDCW"^-1*" h"^-1),
  scales = list(alternating = FALSE),
  panel=function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.errbars(x, y, ewidth = 0, ...)
    panel.key(..., which.panel = 2, corner = c(0.9, 0.9))
  }
)

print(plot_CO2_em, split = c(1,1,1,2), more = TRUE)
print(plot_O2_up, split = c(1,2,1,2))
```
```{r, include = FALSE}
# silently export svg figures
svg("../figures/figure_CO2_O2.svg", width = 5, height = 5.5)
print(plot_CO2_em, split = c(1,1,1,2), more = TRUE)
print(plot_O2_up, split = c(1,2,1,2))
dev.off()
```


## High throughput imaging of chemostat cultivated cells

The same samples as were used to determine bulk PHB content using nile red fluorescence spectroscopy were also subjected to high throughput imaging.
A Nikon microscope was used to acquire up to 5-10 images per replicate and condition. Phase contrast images and Texas Red fluorescence were acquired to identify cell outline/shape and Nile red fluorescnce from PHB staining, respectively. A custom Cell Profiler pipeline was used to automatically identfy cells, filter out non-cell objects by shape/size, and quantify fluorescence.

First we load a processed data frame combining data of fructose and ammonium limitation.

```{r, message = FALSE}
df_imaging <- read_csv("../data/input/20190728_chemostat_imaging.csv")
```

Now we can plot the length of the cells and the volume calculated from the major axis length, and from the assumption that cells are rod/cylinder shaped objects. More precisely, cell volume using the volume of a cylinder as model was determined using the height h, (longest/major axis of our cell image), and its radius r (half of the minor axis/diameter of our cells). Then, volume is calculated as V = pi x r^2 x h.

```{r, fig.width = 5, fig.height = 5.5}
# see graph pars "box.rectangle" for the box, "box.umbrella" for whiskers
cell_length <- xyplot(MajorAxisLength_um ~ factor(Metadata_mu) | limitation, 
  df_imaging %>% filter(strain == "WT (PHB+)"),
  par.settings = custom.colorblind(),
  scales = list(alternating = FALSE),
  xlab = expression("µ [h"^-1*"]"),
  ylab = expression("major axis length [µm]"),
  lwd = 3, ylim = c(-0.5, 6.5),
  as.table = TRUE, between = list(x = 0.5, y = 0.5),
  panel=function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.bwplot(x, y, horizontal = FALSE, pch = "|", 
      notch = FALSE, do.out = FALSE, ...)
    n_cells = tapply(y, x, length)
    panel.text(labels = paste("n > ", min(n_cells, na.rm = TRUE)-1), x = 2, y = 5.5)
    #lm(y ~ x, data.frame(x = x, y= y)) %>% summary %>% print
  }
)

cell_volume <- xyplot(cell_volume_um3 ~ factor(Metadata_mu) | limitation, 
  df_imaging %>% filter(strain == "WT (PHB+)"),
  par.settings = custom.colorblind(),
  scales = list(alternating = FALSE),
  xlab = expression("µ [h"^-1*"]"),
  ylab = expression("cell volume [µm"^3*"]"),
  lwd = 3, ylim = c(-3, 33),
  as.table = TRUE, between = list(x = 0.5, y = 0.5),
  panel=function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.bwplot(x, y, horizontal = FALSE, pch = "|", 
      notch = FALSE, do.out = FALSE, ...)
    n_cells = tapply(y, x, length)
    panel.text(labels = paste("n > ", min(n_cells, na.rm = TRUE)-1), x = 2, y = 28)
    #lm(y ~ x, data.frame(x = x, y= y)) %>% summary %>% print
  }
)

print(cell_length, split = c(1,1,1,2), more = TRUE)
print(cell_volume, split = c(1,2,1,2))
```


```{r, include = FALSE}
# silently export svg figures
svg("../figures/figure_imaging.svg", width = 5, height = 5.5)
print(cell_length, split = c(1,1,1,2), more = TRUE)
print(cell_volume, split = c(1,2,1,2))
dev.off()
```


```{r, include = FALSE}
# for now excluded because growth rate 0.05 h^-1 samples were not quantified correctly
# (problem with sample preparation)
# here we additionally filter one erroneous replicate out
xyplot(Intensity_MedianIntensity_TxRed_0_resize ~ factor(Metadata_mu) | strain * limitation, 
  df_imaging %>% filter(!(Metadata_mu == 0.25 & Metadata_reactor == 1)),
  par.settings = custom.colorblind(),
  scales = list(alternating = FALSE),
  xlab = "", lwd = 3, #ylim = c(-3, 33),
  as.table = TRUE, between = list(x = 0.5, y = 0.5),
  panel=function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.bwplot(x, y, horizontal = FALSE, pch = "|", 
      notch = FALSE, do.out = FALSE, ...)
    n_cells = tapply(y, x, length)
    panel.text(labels = paste("n > ", min(n_cells, na.rm = TRUE)-1), x = 2, y = 0.9)
  }
)
```



