---
title: "CRISPRi library additional tests"
output: html_notebook
---


## Description

The purpose of this R markdown notebook is to provide a trackable, reproducible,
and comprehensive documentation of data processing steps for experiments carried 
out using a CRISPRi sgRNA library in the  cyanobacterium _Synechocystis_ sp. PCC6803.

This is an additional part of the pipeline where other tracks are explored by
blending the fitness results of the CRISPRi library with protein abundance
or protein variability. The additional data set used here is a proteomics study 
from [Jahn et al., Cell Reports, 2018](https://linkinghub.elsevier.com/retrieve/pii/S2211124718314852).
The idea is that gene/protein essentiality could correlate with variability in
protein abundance. One scenario would be that 'core proteins' are both essential and
less variable in abundance as they are required under all possible conditions.
The 'core proteome' (around 400 proteins in _E. coli_) was reported to  be mostly
allosterically regulated instead of transcriptionally.


```{r, results = 'hide', message = FALSE}
# LOAD PACKAGES

library(lattice)
library(latticeExtra)
library(grid)
library(tidyverse)
library(Rtools)
```


## Correlation between gene fitness and gene expression variability

Preprocessed data in the form of reads per gene was summarized and annotated 
in pipeline 1. It includes mean read fraction, log2 FC and p-value calculated 
using DESeq2. Genome annotation was added based on uniprot and cyanobase annotation
of the reference genome for _Synechocystis_ sp. PCC6803. 

----------

#### Loading and merging proteomics data

```{r, message = FALSE, warning = FALSE}
# load CRISPRi library data
load(file = "../processed_data/CRISPRi_library_df_cluster.Rdata")

# exclude ncRNAs from analysis
df <- df %>% filter(!grepl("^Entry", sgRNA))

# load MS data and combine with library;
# matching conditions are mapped to each other (L300 and L100 conditions);
# the MS data does not contain LD cycle

df <- read_csv("../../../../../Experiments/20171204_MS_CO2/CO2_light/analysis/diffacto_weightedsum/df_long_massfrac.csv") %>%
  
  # load mass fraction data from Cell Reports paper
  filter(sample == "light", light %in% c(300, 100)) %>%
  
  # change sample annotation from [100, 300] to [L100, L300]
  mutate(condition = recode(light, `300` = "L300", `100` = "L100")) %>%
  
  # rename a column
  rename(locus = protein) %>%
  
  # select only the interesting columns
  select(locus, condition, mean.mass.fraction.norm, ANOVA, ANOVA.adj) %>%
  
  # merge with df
  left_join(df, .) %>%
  
  # filter for L100 and L300 only
  filter(condition %in% c("L100", "L300"))
  
# test if mass fraction sums up to _around 1_ (also depends on intersection
# between proteomics and CRISPRi library gene sets)
df %>% group_by(condition, timepoint, induction, sgRNA_index) %>% 
  summarise(sum(mean.mass.fraction.norm, na.rm = TRUE)) %>% head


# # optionally reduce data to only the most 
# df <- df %>% group_by(condition, induction, sgRNA_short) %>%
# summarise(
#   fitness_score = min(fitness_score),
#   ANOVA.adj = min(ANOVA.adj)
# )

```

----------

#### Correlation between change over light conditions and essentiality

Different metrics could be used to determine a the extent of a protein changing 
abundance over light conditions. The MS dataset contains an ANOVA p-value that is 
lower (more significant) the more a protein changes abudnance over light intensity/
growth rate (5 light intensities were originally tested: L60, L100, L200, L300 
and L1000). The p-value obtained from ANOVA is the same for all conditions. It is
therefore not neccessary to include all possible light conditions. If all sgRNAs 
are mapped to their corresponding proteins (2 sgRNAs per protein), no correlation is
visible. Also not, if optionally only the more depleted of the two sgRNAs per protein
are selected.

```{r, fig.align = 'center', fig.width = 6, fig.height = 3.5}
# filter dataset for 1 time point only (fitness score is identical
# for all time points of 1 condition)
df %>% filter(timepoint == 0) %>%
  
  xyplot(-log10(ANOVA.adj) ~ fitness_score | factor(condition, unique(condition)), .,
  as.table = TRUE, groups = induction, 
  between = list(x = 0.5, y = 0.5),
  par.settings = custom.lattice, pch = 19, cex = 0.3,
  scales = list(alternating = FALSE),
  #xlim = c(-9, 4), ylim = c(-9, 4),
  xlab = "fitness", ylab = "ANOVA -log10 p-value",
  panel = function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.xyplot(x, y, ... )
    panel.key(c("induced", "uninduced"), pch = 19, corner = c(0.1, 0.9))
  }
)
```

However it is possible to refine the approach by selecting subsets of sgRNAs or 
proteins. In the following step, proteins are binned by ANOVA log10 -p-value,
and fitness score distribution evaluated as box and whisker plot. Here, a trend
can be seen in the form of more significantly regulated genes (higher -log10 p-value)
having reduced median fitness.

```{r, fig.align = 'center', fig.width = 5, fig.height = 4}
# customize plotting colors a bit
custom.lattice <- custom.lattice()
custom.lattice$box.rectangle$col = rep(c("#CC476B", "#00526D"), c(3,2))
custom.lattice$box.umbrella$col = rep(c("#CC476B", "#00526D"), c(3,2))
custom.lattice$box.rectangle$lwd = 1.5
custom.lattice$box.umbrella$lwd = 1.5


# filter dataset for 1 time point only (fitness score is identical
# for all time points of 1 condition)
plot_fitness_ANOVA_bw <- df %>% filter(timepoint == 0) %>%
  
  # bin into groups by -log10 p-value
  mutate(ANOVA_bin = cut(ANOVA.adj, breaks = c(1,0.1,0.05,0.01,0.001,0))) %>%
  
  xyplot(fitness_score ~ factor(ANOVA_bin) | factor(condition) %>% 
      paste0(" - ", recode(induction, i = "induced", u = "uninduced")), .,
    as.table = TRUE, between = list(x = 0.5, y = 0.5),
    par.settings = custom.lattice, lwd = 1.5, 
    scales = list(alternating = FALSE, x = list(rot = 30, cex = 0.7)),
    xlab = "ANOVA -log10 p-value", ylab = "fitness",
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.violin(x, y, horizontal = FALSE, border = NA, col = grey(0.7, 0.5))
      panel.bwplot(x, y, horizontal = FALSE, pch = "|", do.out = FALSE, ... )
      panel.pvalue(x, y, fixed_pos = -1, std = 5, 
        col = rep(c("#CC476B", "#00526D"), c(3,2))
      )
    }
  )

print(plot_fitness_ANOVA_bw)
```


As a follow-up, the most regulated proteins were selected and the fitness of the 
corresponding sgRNAs was plotted as barchart (groups in red from above, 
$p_{value} \leq 0.05$). 



```{r, fig.align = 'center', fig.width = 5, fig.height = 4}
# customize plotting colors a bit
custom.lattice$box.rectangle$col = "#CC476B"#"#00526D"
custom.lattice$box.umbrella$col = "#CC476B"#"#00526D"

# filter dataset for 1 time point only (fitness score is identical
# for all time points of 1 condition)
plot_fitness_ANOVA_pw <- df %>% 
  
  # filter also for most regulated genes = p-value <= 0.01
  filter(timepoint == 0, induction == "i", ANOVA.adj <= 0.05) %>% 
  
  # arrange genes by mean fitness over pathway
  group_by(Process.abbr) %>% 
  mutate(
    mean_fitness = mean(fitness_score, na.rm = TRUE),
    n_sgRNAs = length(unique(sgRNA))) %>%
  arrange(mean_fitness) %>%
  
  # plot fitness per pathway
  xyplot(fitness_score ~ paste0(Process.abbr, " (", n_sgRNAs, ")") %>% factor(., unique(.)) | 
      paste0(condition, " - induced"), .,
    as.table = TRUE, between = list(x = 0.5, y = 0.5),
    par.settings = custom.lattice, lwd = 1.5, layout = c(1,2),
    scales = list(alternating = FALSE, x = list(rot = 30, cex = 0.7)),
    xlab = "cyanobase pathway", ylab = "fitness",
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.violin(x, y, horizontal = FALSE, border = NA, col = grey(0.7, 0.5),... )
      panel.bwplot(x, y, horizontal = FALSE, pch = "|", do.out = FALSE, ... )
    }
  )

print(plot_fitness_ANOVA_pw)
```

```{r, results = 'hide', include = FALSE}
svg("../figures/supplemental/Supplemental_figure_prot_regulation.svg", width = 5, height = 7)
print(plot_fitness_ANOVA_bw, position = c(0, 0.48, 1, 1.04), more = TRUE)
print(plot_fitness_ANOVA_pw, position = c(0, 0, 1, 0.53))
grid.text(label = c("A", "B", "C"), x = c(0.03, 0.03), y = c(0.98, 0.5), gp = gpar(cex = 1.3))
dev.off()
```

----------

**Supplementary Figure. Highly regulated proteins have lower average fitness.**
**A.** Fitness score of sgRNAs broken down by protein variability. Protein variability
was estimated using a proteomics study with measurement of protein abundance over several
light conditions (Jahn et al., Cell Reports, 2018). Here, the adjusted p-value from 
ANOVA over all light conditions was used as a metric for variability. For this comparison,
every sgRNA was mapped to its corresponding protein. Red, proteins with $p_{value} \leq 0.05$).
Blue, proteins with $p_{value} > 0.05$).
**B.** Fitness score of sgRNAs associated with most significantly changing proteins
($p_{value} \leq 0.05$, red in (A), broken down by cyanobase pathways.
In brackets, number of unique sgRNAs per pathway.

----------
