---
title: "CRISPRi library data processing pipeline"
output:
  html_notebook:
    toc: yes
---


## Description

The purpose of this R markdown notebook is to provide a trackable, reproducible,
and comprehensive documentation of data processing steps for experiments carried 
out using a CRISPRi sgRNA library in the  cyanobacterium _Synechocystis_ sp. PCC6803.

This is the first pipeline of two, focusing on loading and processing the obtained
highly parallel sequencing data from growth competiion experiments. The experimental 
details will be published in a corresponding paper. The complete data set consists
of several independent experiments sequenced at different time points that are
merged into one complete data frame eventually. The first steps of this pipeline
are reading the (preprocessed) data tables for low light (L100), high light (L300),
LD (light-dark-cycle, a.k.a. day-night-cycle), sodium lactate tolerance (LAC), 
and a sodium chloride control (NACL). This is followed by adding some statistical 
metrics and a basic quality control regarding read counts, and number of identified genes.


```{r, results = 'hide'}
# LOAD PACKAGES

library(lattice)
library(latticeExtra)
library(DESeq2)
library(tidyverse)
library(Rtools)
library(DescTools)
```


## Step 1: Preprocessing of data

Preprocessed data in the form of reads per gene were converted to fraction of 
total reads by dividing each read number by total number of reads, per condition 
and replicate. 

```{r, results = 'hide'}
df_raw <- list(
  L100 = read_tsv("../raw_data/190428_countings_long_format_MERGED_LL.tab"),
  L300 = read_tsv("../raw_data/190428_countings_long_format_MERGED_HL.tab"),
  LD = read_tsv("../raw_data/190217_countings_long_format_MERGED_DN.tab"),
  LAC = read_tsv("../raw_data/190430_countings_long_format_lactate_tolerance.tab"),
  NACL = read_tsv("../raw_data/190508_countings_long_format_NaCl_control.tab")
  ) %>% bind_rows(.id = "condition") %>%
  
  # add class and rename some columns
  dplyr::rename(replicate = sample) %>%
  
  # make separate group for antisense RNA loci
  mutate(type = ifelse(grepl("^Entry", sgRNA), "ncRNA", "gene")) %>%
  
  # change timepoint to numeric value
  mutate(timepoint = as.numeric(gsub("[hd]", "", timepoint))) %>%
  mutate(timepoint = plyr::mapvalues(timepoint, c(24,48), c(1,2))) %>%
  
  # change some replicate indices from 5-8 to 1-4
  mutate(replicate = plyr::mapvalues(replicate, 5:8, 1:4)) %>%
  
  # grouping for condition, time, induction
  group_by(type, condition, timepoint, induction, replicate) %>%
  
  # and add read fraction for each combination (cond, time, ind, repl)
  mutate(read.fraction = readcount/sum(readcount))
```

The correct calculation of read fraction is checked by the following steps, 
including summarizing unique sgRNAs and mean read count per condition.

```{r}
# check individual sgRNAs per sample, should be around 7000 (2 per gene)
summarise(df_raw, length = length(read.fraction)) %>% head
# check that every condition sums to 1
summarise(df_raw, sumFC = sum(read.fraction)) %>% head
# altogether are 149 samples (6xL300, 4xL100, 6xLD, 3xNACL, 3xLAC, most of them +/- induction, 4 repl)
# (times 2 for differentiation between gene and ncRNA loci)
sum(df_raw$read.fraction)
# check also mean raw read count per sample
summarise(df_raw, readcount = median(readcount)) %>% head
```

One particular control condition, the lactate-tolerance sample for time point zero, 
induced (`LAC_0d_induced`), was not sequenced individually as it is identical
with the non-induced counter part of the same condition (`LAC_0d_uninduced`). 
For calculation of fold change metric, the missing condition was simply copied 
from the present control.

```{r}
df_raw <- filter(df_raw, condition == "LAC" & timepoint == 0 & induction == "u") %>%
  ungroup %>% mutate(induction = "i") %>%
  bind_rows(., df_raw)
```

A preliminary analysis of raw read counts for all different conditions, timepoints, 
and replicates was conducted. For the sake of simplicity, the time points 12 and
and 30 days (sodium chloride control, condition `NACL`) were mapped to 
16 and 32 days to correspond to time points of all other samples. This does not
influence calculation of generation time or fitness score, and was done only to
display the data.

```{r, fig.width = 10, fig.height = 7}
plot_median_read_count <- df_raw %>% 
  
  # focus on genes
  filter(type == "gene") %>%
  
  # first group by all possible conditions
  group_by(condition, timepoint, induction, replicate) %>%
  
  # determine read count
  summarise(median_read_count = median(readcount)) %>% ungroup %>%
  mutate(timepoint = plyr::mapvalues(timepoint, c(12, 30), c(16, 32))) %>%
  
  # and plot
  xyplot(median_read_count ~ factor(timepoint) | induction * condition, .,
    xlab = "time [d]", ylab = "median read count",
    as.table = TRUE, par.settings = custom.lattice, 
    scales = list(alternating = FALSE),
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      if (panel.number() %in% 1:9) {
        panel.barplot(x, y, ewidth = 0.3, ...)
      }
    }
  )

# plot also number of individual sgRNAs per sample
plot_unique_sgRNA <- df_raw %>% 
  
  # focus on genes
  filter(type == "gene") %>%
  
  # first group by all possible conditions
  group_by(condition, timepoint, induction, replicate) %>%
  
  # determine read count
  summarise(unique_sgRNAs = length(readcount)) %>% ungroup %>%
  mutate(timepoint = plyr::mapvalues(timepoint, c(12, 30), c(16, 32))) %>%

  xyplot(unique_sgRNAs ~ factor(timepoint) | induction *  condition, .,
    pch = 19, type = c("p"),
    xlab = "time [d]", ylab = "unique sgRNAs",
    as.table = TRUE, par.settings = custom.lattice, 
    scales = list(alternating = FALSE),
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      if (panel.number() %in% 1:9) {
        panel.barplot(x, y, ewidth = 0.3, ...)
      }
    }
  )

print(plot_median_read_count, split = c(1,1,2,1), more = TRUE)
print(plot_unique_sgRNA, split = c(2,1,2,1))
```

```{r, results = 'hide', include = FALSE}
svg("../figures/supplemental/CRISPRi_library_read_counts.svg", width = 10, height = 7)
print(plot_median_read_count, split = c(1,1,2,1), more = TRUE)
print(plot_unique_sgRNA, split = c(2,1,2,1))
dev.off()
```
-----------

**Supplementary Figure.**
**A**. Median read count per condition. Bars and whiskers correspond to mean and standard deviation of 4 biological replicates. 
**B**. Coverage of samples in terms of quantified unique sgRNAs. Bars and whiskers correspond to mean and standard deviation of 4 biological replicates. Abbreviations: L100 - light with 100 µmol/m2\*s, L300 - light with 100 µmol/m2\*s, LD - light-dark-cycle, LAC - addition of lactate, NACL - addition of sodium chloride.

-----------


## Step 2: DESEQ2 analysis for fold change and significance testing

To obtain fold changes between conditions and a significance metric, the R package
DESeq2 was used. DESeq2 uses a negative binomial distribution of read counts to make more
accurate estimation of fold change and significance per gene. DESeq2 reference:

*Love, M.I., Huber, W., Anders, S. (2014) Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biology, 15:550. https://doi.org/10.1186/s13059-014-0550-8*

#### 1. Read count matrix

The data frame was reshaped for DESeq2 into a 'counts matrix' with genes as rows 
and samples (conditions) as columns.

```{r}
counts <- select(df_raw, condition, induction, timepoint, replicate, sgRNA, readcount) %>%
  
  # paste metadata columns together
  unite(condition, condition, induction, timepoint, replicate) %>%
  
  # spread condition over columns and sgRNAs over rows
  spread(condition, readcount) %>%
  
  # remove sgRNA column, replace NA with 0
  mutate_at(vars(-1), function(x) coalesce(x, 0)) %>%
  
  # add row_names from column 
  column_to_rownames("sgRNA")

head(counts)
```
  

#### 2. Meta data

A meta data frame with experimental annotation is created. This is used to carry 
out the actual DESeq2 analysis by 'contrasting' (comparing) selected conditions 
to each other.


```{r}
# The meta data frame with experiment annotation
samples <- data.frame(
  all_conditions = colnames(counts),
  # Order must be same as Reads columns!
  row.names = colnames(counts) 
  ) %>%
  
  # split the metadata into single variables
  separate(all_conditions, into = c("condition", "induction", "timepoint", "replicate")) %>%
  
  # make combinations of condition, induction and time to form super-groups,
  # then test each time point against each other within the group
  unite(group, condition, induction, timepoint, remove = FALSE)

head(samples)
```

#### 3. Perform DESeq2 analysis

It is possible to use complex experimental designs and specify those
using a formula interphase. However, the manual recommends
to merge combinations of different factors into super-groups
and then do pairwise comparison of the groups one is interested in.

**WARNING: This step is computation-intense and can take several hours**

```{r, eval = FALSE}
  DES_result <- DESeqDataSetFromMatrix(
    countData = counts,
    colData = samples,
    design = ~ group) %>%
    DESeq
  
  
  # Have alook at the combinations that were tested
  resultsNames(DES_result)[1:10]
```

Next, results from DESeq2 models are extracted. The data set includes
5 conditions x 2 inductions x 3/4/6 timepoints = 38 combinatiions in total.
The syntax to call DESeq2's `results(...)` function is to use one pair of 
contrasts `contrast("variable", "level1", "level2")`. To automate this, 
a list of condition and reference pairs is set up. As mentioned before, DESEq2 
has more advanced ways of constructing contrasts but the syntax is more difficult.

```{r, results = 'hide', eval = FALSE}
# list of contrasts/combinations
combinations <- samples$group %>% unique %>% 
    grep(pattern = "_([1-9]*|30)$", value = TRUE) %>%
  list(testCond = ., refCond = gsub("_([1-9]*|30)$", "_0", .))


# loop through all possible combinations
DES_result_table = lapply(1:length(combinations$testCond),
  
  function (i) {
    results(DES_result, 
      contrast = c("group", combinations$testCond[[i]], combinations$refCond[[i]])) %>%
    as.data.frame %>% 
    mutate(
      sgRNA = rownames(.), 
      group = combinations$testCond[[i]]
    ) %>% as_tibble %>%
    separate(group, into = c("condition", "induction", "timepoint")) %>%
    mutate(timepoint = as.numeric(timepoint))
  }
  
) %>% bind_rows
```

----------
Alternatively, it is possible to load a precumputed DESeq2 results table. This avoids
re-running the entire DESeq2 analysis only for another part of the pipeline has
changed.

```{r, results = 'hide', eval = FALSE}
# load precomputed DESeq2 result in form of Rdata object
load("../processed_data/CRISPRi_library_DESeq_result_table.Rdata")
```


## Step 4: Merge with DESeq2 results and summarize data

The following steps condense the raw data frame by averaging replicates,
merging the data frame with the result from DESeq2 analysis, and adding different
statistical metrics.

```{r}
# change the grouping to condition, timepoint, induction, sgRNA
df <- group_by(df_raw, condition, timepoint, induction, sgRNA) %>% 
  
  # summarise replicates by mean and stdev
  summarise(
    read.fraction.mean = mean(read.fraction, na.rm = TRUE),
    read.fraction.stdev = sd(read.fraction, na.rm = TRUE)) %>%
  
  # new grouping for condition, induction, sgRNA BUT NOT timepoint
  group_by(condition, induction, sgRNA) %>%
  
  # determine FC over time by comparing each sgRNA at each condition
  mutate(FC = read.fraction.mean / read.fraction.mean[1])
  
# test if mean FC == 1 for all 0d time points and != 1 for all others
df %>% group_by(condition, timepoint, induction) %>%
  summarise(meanFC = mean(FC, na.rm = TRUE)) %>% head

# merge the dataframe with the result from DESeq2
# this is not trivial because sgRNA and condition have different length
# (DESeq contains all possible combinations of sgRNA and condition, df not)
# a 'left_join' will keep the structure of df
df <- left_join(df, DES_result_table) %>% ungroup
```


## Step 5: Annotation of data

The data frame has to be completed with `NA` for sgRNAs missing in some 
combinations of conditions. Here, `tidyr::complete` was used to fill these gaps. 
The syntax is to indicate first the variable that is to be completed, and then the 
variables that represent conditions to be completed. 
All other variables (like measurements) are filled with `NA`. 

Different annotation columns are added to the master data frame, including a short sgRNA
identifier (excluding the position on the gene), an sgRNA index (1 or 2), and the
customized genome annotation from Uniprot/KEGG.

```{r, results = 'hide'}
df <- df %>% complete(sgRNA, nesting(condition, timepoint, induction)) %>%

  # prepare merging by cleaning IDs from results table
  # first remove sgRNA position number from gene IDs
  mutate(sgRNA_short = 
  gsub("(\\_| )[0-9]+(\\_MULTI)?$", "", sgRNA)) %>%
  
  # add sgRNA index number (1 or 2)
  group_by(condition, timepoint, induction) %>%
  mutate(sgRNA_index = duplicated(sgRNA_short) %>% as.numeric %>% {. + 1}) %>%

  # map trivial names to LocusTags using a manually curated list
  left_join(
    read_tsv("../raw_data/gene_protein_table.txt") %>% 
      filter(!duplicated(gene)), 
    by = c("sgRNA_short" = "gene")) %>%
  
  # one more join, this time with the annotation db
  # containing uniprot and categorial (KEGG Brite) data
  left_join(read_csv("../raw_data/Synechocystis_PCC6803_genome_annotation_20190614.csv")[c(1,4:6,11,20,22,27:32)], 
    by = c("locus" = "GeneID"))

```

Check that number of detected sgRNAs is the same for all conditions
after completing.
  
```{r}
  df %>% summarise(length(unique(sgRNA))) %>% head
```

----------

The number of generations for each experiment and time point was calculated 
based on known cultivation time and average growth rate in the turbidostat.
A fitness score was defined as the area under/over the curve for log2 fold change
over number of generations. An enrichment will result in a positive score, a depletion 
in a negative score. The fitness score is normalized to the maximum number of 
generations for a particular condition, and is therefore independent of the duration
of a particular cultivation. The fitness score will be lower for faster depleting genes
and higher for slowly depleting genes.

```{r, results = 'hide'}
# using left join with table
df <- left_join(df, read_csv("../raw_data/growth_rate_library.csv")) %>%

  # first fill up missing (NA) log2FC values for the initial 0d time point
  group_by(condition, induction, sgRNA) %>%
  mutate(log2FoldChange = ifelse(FC == 1, 0, log2FoldChange)) %>%
  
  # determine fitness score using DescTools::AUC()
  mutate(fitness_score = AUC(generations, log2FoldChange)/max(generations)) %>%
  
  # make temporary change of NACL time points 12 and 30 days to 16 and 32, 
  # to ease comparisons. That does not effect the fitness score that was calculated
  # with the accurate time points, nor the number of generations
  ungroup %>% mutate(timepoint = plyr::mapvalues(timepoint, c(12, 30), c(16, 32)))

```

```{r}
  # check how many sgRNAs have valid scores
  df %>% group_by(condition, induction) %>% 
    summarise(total = length(fitness_score), NAs = sum(is.na(fitness_score)))
```


## Step 6: Quality control

The log2 FC that DESeq2 determines takes measures of read density or distribution 
into account that a manually determined log2 FC does not. A good control of how 
different a manual log2 FC is from the DESeq2 log2 FC is to simply plot one 
against the other. There are only minor differences.

```{r, fig.width = 8, fig.height = 7}
plot_DESeq2_vs_namual_FC <- df %>% 
  
  # plot only genes and remove time point zero
  filter(!grepl("Entry", sgRNA),
    !is.na(log2FoldChange), timepoint != 0) %>%
  
  xyplot(log2FoldChange ~ log2(FC) | factor(timepoint) * 
      factor(condition, c("L100", "L300", "LD", "LAC", "NACL")), ., 
    as.table = TRUE, groups = induction,
    par.settings = custom.lattice, 
    xlab = "manual log2FC", ylab = "DESeq2 log2FC",
    pch = 19, cex = 0.3, 
    scales = list(alternating = FALSE),
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.xyplot(x, y, ...)
      panel.key(c("induced", "uninduced"), pch = 19, corner = c(0.1, 0.9))
    }
  )

print(plot_DESeq2_vs_namual_FC)
```


```{r, results = 'hide', include = FALSE}
png("../figures/supplemental/CRISPRi_library_DESeq2_vs_manual_FC.png", width = 1000, height = 800, res = 100)
print(plot_DESeq2_vs_namual_FC)
dev.off()
```


----------

A second check of DESeq2's calculation of log2 FC and p-value is to visualize both
metrics in a volcanoplot. The log2 FC is usually on the x-axis and the negative log10 
p-value on the y-axis, showing the most significantly _and_ strongly changed genes 
(here sgRNAs) in the upper right and left corner.

```{r, fig.width = 8, fig.height = 7}
plot_volcanoplot <- df %>%
  
  # plot only genes and remove time point zero
  filter(!grepl("Entry", sgRNA),
    !is.na(log2FoldChange), timepoint != 0) %>%
  
  xyplot(-log10(padj) ~ log2FoldChange | factor(timepoint) * 
      factor(condition, c("L100", "L300", "LD", "LAC", "NACL")), ., 
    as.table = TRUE, groups = induction,
    xlab = "log2 FC", ylab = "-log10(p value)",
    par.settings = custom.lattice, 
    cex = 0.2, pch=19,
    scales = list(alternating = FALSE),
    ylim = c(- 20, 200), xlim = c(-10, 10),
    panel = function(x, y, ...) {
      panel.grid(h = -1, v = -1, col = grey(0.9))
      panel.xyplot(x, y, ...)
      panel.key(c("induced", "uninduced"), pch = 19, corner = c(0.1, 0.9))
    }
  )

print(plot_volcanoplot)
```


```{r, results = 'hide', include = FALSE}
png("../figures/supplemental/CRISPRi_library_volcanoplot.png", width = 1000, height = 800, res = 100)
print(plot_volcanoplot)
dev.off()
```

-----------

**Supplementary Figure.**
Volcanoplot of log2 FC versus negative log10 transformed p-value as result of DESeq2 analysis. Each panel shows one condition and time point with overlay of induced and uninduced samples (color-coded in blue and red, respectively). Induction of sgRNA expression was performed by adding IPTG.
Abbreviations: L100 - light with 100 µmol/m2\*s, L300 - light with 100 µmol/m2\*s, LD - light-dark-cycle, LAC - addition of lactate, NACL - addition of sodium chloride.

-----------

The last step is to save the DESeq2 result (computatioanlly intense) and the 
processed data frame to file and continue with the actual analysis of sgRNA enrichment
and depletion in the next part of the pipeline.


```{r}
save(df, file = "../processed_data/CRISPRi_library_df_annotated.Rdata")
save(DES_result_table, file = "../processed_data/CRISPRi_library_DESeq_result_table.Rdata")
```


